---
title: "Chapter 20"
code-overflow: wrap
---

Chapter 20 of *R for Data Science* covers how to work with spreadsheets from Excel and Google Sheets. To get started, I'll load a few essential packages:

```{r}
pacman::p_load(tidyverse, readxl, writexl)
```

```{r}
#write csv displayed in R4DS, for use in subsequent problems

survey <- read_csv("survey_id, n_pets\n1, 0\n2, 1\n3, N/A\n4, two\n5, 2\n6, NA", quote = '', show_col_types = FALSE)

survey

write_xlsx(survey, path = "./survey.xlsx")
```

# 20.2.9 Exercises

## Exercise 20.2.9.1

**In an Excel file, create the following dataset and save it as survey.xlsx. Then, read it into R, with survey_id as a character variable and n_pets as a numerical variable.**

```{r}
#n_pets must start as character type to preserve the value "two" in the 4th row. We can replace this with "2" and then parse this column to create a numeric column type

read_excel(path = "./survey.xlsx", col_types = "text") %>%
  mutate(
    n_pets = parse_number(if_else(n_pets == "two", "2", as.character(n_pets))))
```

## Exercise 20.2.9.2

**In another Excel file, create the following dataset and save it as roster.xlsx**

```{r}
roster <- read_csv("group, subgroup, id\n1, A, 1\n,,2\n,,3\n, B, 4\n,, 5\n, , 6\n,, 7\n2, A, 8\n,, 9\n, B, 10\n,, 11\n,, 12", show_col_types = FALSE)

roster

write_xlsx(roster, path = "./roster.xlsx")

read_excel("./roster.xlsx") %>% 
  fill(everything())
```
